\vspace{-2.0cm}The effect of nitrogen application time on infection and development of the Wheat Streak Mosaic Virus (WSMV)

### 2015-10-01 draft1leslieandkenny.Rnw
Wrote out glm model and ran spellcheck.

Nar conducted this experiment in a MSU field at the base of the Bridger Mountains. He chose one area in the field and divided it into six $31.5$ by $27.5$ meter blocks. The blocks were then divided into five rows, with $5$ meters of space between each row. Each row was randomly assigned to a variety of Winter Wheat with separate randomizations in each block. Three of the chosen varieties are known to be resistant to WSMV (SNMS, TAM 112, and MACE). The pronghorn variety (PRHG) is known to be susceptible to WSMV. The last variety, Yellowstone (YLST), is a common variety in Montana, and its susceptibility to WSMV is unknown. \\

@@

\noindent The sampling was all done in one day on two occasions. The first sampling date occured at the end of May when the plants were in the tillering stage, and the second sampling date occured in early July when plants were in the flowering stage. In the first year of the study, $2013-2014$, all plots were sampled on both dates. In the second year of the study, only plots with susceptible varieties were sampled on the first date, and all plots were sampled on the second date The advice given in this report is only for analysis of the data collected on the second date in July. After the leaves were picked in the field, they were sent to the lab and screened for the virus via the ELISA procedure. The data are collected and organized in a spreadsheet for the first year. The second year leaves are currently stored in ziploc bags in the freezer and are waiting to be analyzed in the lab. The researchers do not expect this waiting period to affect their ability to detect the virus. \\

@@

\subsection{Binomial Generalized Linear Mixed Model}

\noindent Because the response variable is a Binomial count, and because
the design includes random assignment of variety to rows within blocks,
a binomial generalized linear mixed model is appropriate. This model
will estimate the probability of infection for each
\verb|variety:treatment:status| combination while controlling for
differences between blocks and between rows in the same block.

\noindent The model is
$$
y_i\sim\mathrm{Binomial}(30, p_i),
$$
$$
\mathrm{logit}(p_i)=\mathbf{x}_i\mathbf{\beta}+b_{0,j[i]}+b_{1,j[i],k[i]}
$$
where
\begin{itemize}
\item \(p_i\) is the probability of infection in the \(i\)th plot,
\item \(\mathrm{logit}(p_i)=\log\left(\dfrac{p_i}{1-p_i}\right)\) is the
natural logarithm of the odds of infection,
\item \(\mathbf{x_i}\) is a row vector containing indicator variables for
the variety, nitrogen application timing, and inoculation status of
observation \(i\),
\item \(\mathbf{\beta}\) is a column vector of fixed-effect coefficients, and
\item \(b_{0,j}\sim\mathrm{N}(0,\sigma^2_{b_0})\) and
\(b_{1,j,k}\sim\mathrm{N}(0,\sigma^2_{b_1})\) are random effects for block
and row, respectively.
\end{itemize}

\noindent The assumptions for this model are that plants within a plot
are independent and have the same probability of being infected, and
that plots within a row are independent of each other.

\noindent Advantages of the Binomial GLMM include:
\begin{itemize}
\item It estimates and controls for variability between rows and blocks.
\item It avoids confounding variety effects with row effects and correctly
models the rows and nested within blocks.
\item Random effects can explain extra variation beyond what the
fixed-effect Binomial GLM can account for.
\end{itemize}

\noindent Disadvantages of this model include:
\begin{itemize}
\item It must be fit by numerically maximizing the likelihood with an
iterative algorithm. Including a large number of parameters (such as
third-order interaction coefficients) can cause the algorithm to fail
to converge.
\item The domain of the logit function is \((0, 1)\). If a large proportion
of the data are zeros, meaning an infection probability is near 0, the
algorithm will have difficulty estimating the coefficients. In this
situation, estimates may be unreliable or the algorithm may not converge.
\item We are not aware of any tools that can fit a Binomial GLMM with
overdispersion. If there are additional sources of variation that the
model does not account for, the standard error estimates will be too small.
\end{itemize}

@@

\begin{figure}[b]
\caption{Example interaction plot using \texttt{qplot}.}
<<interactionplot,echo=FALSE,message=FALSE>>=
require(ggplot2)
qplot(x = n.trt, y = infected, geom = 'boxplot', color = status,
      facets = .~variety, data = fert.sim1) +
  theme(axis.text.x = element_text(angle = 90))
@
\end{figure}

\noindent We noticed that infection counts of 0 were recored in 67 of the
180 plots in Nar's dataset. This is a large fraction of the data; if the
distribution of these zeros is related to the variables in the model then
it may be impossible to estimate coefficients for factor levels where
infection rates are nearly zero.

@@

\noindent If a treatment has many zeros across all varieties or a variety
has many zeros, Nar should consider whether he has reason to believe
that treatment or variety has qualitative differences from the others
that would cause it to have a much lower infrection rate. These data would
be considered as coming from a different population than the data with
fewer zeros. \textbf{If and only if it is justified}, Nar could omit all
of the data for this treatment or variety when fitting the model. The scope
of inference for this model must then be limited to varieties and
treatments that have a nonzero infection rate. A separate analysis would
be needed for more resistant population.

\begin{figure}[b]
\caption{Example heatmap visualization for identifying patterns of zeros.}
<<heatmap,echo=FALSE,message=FALSE>>=
## Heatmap with meaningful ordering

@@

<<startingmodel,warning=FALSE,results='hide',cache=TRUE>>=
glm3way <- glm(cbind(infected, total) ~ block/variety + variety*n.trt*status2,
               family = quasibinomial, data = fert.sim1)
@

If it is decided that certain observations come from an extremely
resistant population and can be omited from the current analysis.

<<novar1,warning=FALSE,results='hide',cache=TRUE>>=
noVar1 <- glm(cbind(infected, total) ~ block/variety + variety*n.trt*status2,
              family = quasibinomial, data = fert.sim1, subset = variety!='Var1')
@

@@

<<separatemodels,warning=FALSE,results='hide',cache=TRUE>>=
onlyINOC <- glm(cbind(infected, total) ~ block/variety + variety*n.trt,
                family = quasibinomial, data = fert.sim1, subset = status=='INOC')

onlyCNTL <- glm(cbind(infected, total) ~ block/variety + variety*n.trt,
                family = quasibinomial, data = fert.sim1, subset = status=='CNTL')
@

<<models5,echo=FALSE,warning=FALSE,results='hide',cache=TRUE>>=
glm3way2 <- glm(cbind(infected, total) ~ block/variety + variety*n.trt*status2,
                family = binomial, data = fert2)
summary(glm3way2)

glm2way2 <- glm(cbind(infected, total) ~ block/variety + variety*n.trt + variety*status2 + n.trt*status2,
                family = quasibinomial, data = fert2)
summary(glm2way2)

Anova(glm3way2, test.statistic = 'F')
Anova(glm2way2, test.statistic = 'F')
Anova(glm2way2, glm3way2, test.statistic = 'F')
anova(glm2way2, glm3way2)
@

<<nosnms,echo=FALSE,warning=FALSE,results='hide',cache=FALSE>>=
table(fert2$infected==0,fert2$variety)

@@

<<plots3,eval=FALSE,echo=FALSE,warning=FALSE,message=FALSE,out.height='0.5\\linewidth',fig.height=5>>=
par(mfrow = c(1, 2))
plot(resid(noSNMS3way) ~ predict(noSNMS3way), col = variety,
     data = fert2, subset = variety!='SNMS')
plot(resid(noSNMS3way) ~ fitted(noSNMS3way), col = variety,
     data = fert2, subset = variety!='SNMS')
legend('topright', pch = 1, col = 1:5,
       legend = levels(fert2$variety))
@

>>>>

Nar conducted this experiment in a MSU field at the base of the Bridger Mountains. He chose one area in the field and divided it into six $31.5$ by $27.5$ meter blocks. The blocks were then divided into five rows, with $5$ meters of space between each row. Each row was randomly assigned to a variety of Winter Wheat with separate randomizations in each block. Three of the chosen varieties are known to be resistant to WSMV (SNMS, TAM 112, and MACE). The pronghorn variety (PRHG) is known to be susceptible to WSMV. The last variety, Yellowstone (YSTN), is a common variety in Montana, and its susceptibility to WSMV is unknown. \\

@@

\noindent The sampling was all done in one day on two occasions. The first sampling date occurred at the end of May when the plants were in the tillering stage, and the second sampling date occurred in early July when plants were in the flowering stage. In the first year of the study, $2013-2014$, all plots were sampled on both dates. In the second year of the study, only plots with susceptible varieties were sampled on the first date, and all plots were sampled on the second date The advice given in this report is only for analysis of the data collected on the second date in July. After the leaves were picked in the field, they were sent to the lab and screened for the virus via the ELISA procedure. The data are collected and organized in a spreadsheet for the first year. The second year leaves are currently stored in Ziploc bags in the freezer and are waiting to be analyzed in the lab. The researchers do not expect this waiting period to affect their ability to detect the virus. \\

@@

\subsection{Binomial Generalized Linear Model}

\noindent Because the response variable is a Binomial count, and because
the design includes random assignment of variety to rows within blocks,
a Binomial generalized linear model is appropriate. We recommend using
a three-way interaction to estimate the probability of infection for each
variety/treatment/inoculation status combination and a block/variety
interaction to control for differences between blocks and between rows
in the same block.

\noindent The model is
$$
y_i\sim\mathrm{Binomial}(30, p_i),
$$
$$
\mathrm{logit}(p_i)=\mu+\alpha_{b[i]}+\beta_{v[i]}+\gamma_{t[i]}+\delta_{j[i]}
+(\alpha\beta)_{b[i],v[i]}+(\beta\gamma)_{v[i],t[i]}+(\beta\delta)_{v[i],j[i]}
+(\gamma\delta)_{t[i],j[i]}+(\beta\gamma\delta)_{v[i],t[i],j[i]}
$$
where
\begin{itemize}
\item \(p_i\) is the probability of infection in the \(i\)th plot,
\item \(\mathrm{logit}(p_i)=\log\left(\dfrac{p_i}{1-p_i}\right)\) is the
natural logarithm of the odds of infection, and
\item \(b[i]\), \(v[i]\), \(t[i]\), and \(j[i]\) index the block,
variety, nitrogen application timing, and inoculation status of
observation \(i\).
\end{itemize}

\noindent The assumptions for this model are that plants within a plot
are independent and have the same probability of being infected, and
that plots are independent given the other variables in the model.

@@

\begin{figure}
\caption{Example interaction plot using \texttt{qplot}.}
<<interactionplot,echo=FALSE,message=FALSE>>=
require(ggplot2)
qplot(x = n.trt, y = infected, geom = 'boxplot', color = status,
      facets = .~variety, data = fert.sim1) +
  theme(axis.text.x = element_text(angle = 90))
@
\end{figure}

\noindent We noticed that infection counts of 0 were recorded in 67 of the
180 plots in Nar's dataset. This is a large fraction of the data; if the
distribution of these zeros is related to the variables in the model then
it may be impossible to estimate coefficients for factor levels where
infection rates are nearly zero.

@@

\noindent If a treatment has many zeros across all varieties or a variety
has many zeros, Nar should consider whether he has reason to believe
that treatment or variety has qualitative differences from the others
that would cause it to have a much lower infection rate. These data would
be considered as coming from a different population than the data with
fewer zeros. \textbf{If and only if it is justified}, Nar could omit all
of the data for this treatment or variety when fitting the model. The scope
of inference for this model must then be limited to varieties and
treatments that have a nonzero infection rate. A separate analysis would
be needed for more resistant population.

\begin{figure}
\caption{Example heatmap visualization for identifying patterns of zeros.}
<<heatmap,echo=FALSE,message=FALSE>>=
## Heatmap with meaningful ordering

@@

<<startingmodel,warning=FALSE,results='hide',cache=TRUE>>=
glm3way <- glm(cbind(infected, total) ~ block*variety + variety*n.trt*status2,
               family = quasibinomial, data = fert.sim1)
@

If it is decided that certain observations come from an extremely
resistant population and can be omitted from the current analysis.

<<novar1,warning=FALSE,results='hide',cache=TRUE>>=
noVar1 <- glm(cbind(infected, total) ~ block*variety + variety*n.trt*status2,
              family = quasibinomial, data = fert.sim1, subset = variety!='Var1')
@

@@

<<separatemodels,warning=FALSE,results='hide',cache=TRUE>>=
onlyINOC <- glm(cbind(infected, total) ~ block*variety + variety*n.trt,
                family = quasibinomial, data = fert.sim1, subset = status=='INOC')

onlyCNTL <- glm(cbind(infected, total) ~ block*variety + variety*n.trt,
                family = quasibinomial, data = fert.sim1, subset = status=='CNTL')
@

<<models5,echo=FALSE,warning=FALSE,message=FALSE,results='hide',cache=TRUE>>=
glm3way2 <- glm(cbind(infected, total) ~ block*variety + variety*n.trt*status2,
                family = binomial, data = fert2)
summary(glm3way2)

glm2way2 <- glm(cbind(infected, total) ~ block*variety + (variety+n.trt+status2)^2,
                family = quasibinomial, data = fert2)
summary(glm2way2)

Anova(glm3way2, test.statistic = 'F')
Anova(glm2way2, test.statistic = 'F')
anova(glm2way2, glm3way2)
@

<<nosnms,echo=FALSE,warning=FALSE,message=FALSE,results='hide',cache=FALSE>>=
table(fert2$infected==0,fert2$variety)

@@

<<plots3,eval=FALSE,echo=FALSE,warning=FALSE,message=FALSE>>=
par(mfrow = c(2, 2))
plot(resid(glm3way2) ~ predict(glm3way2), col = variety, data = fert2)
plot(resid(glm3way2) ~ fitted(glm3way2), col = variety, data = fert2)
plot(resid(noSNMS3way) ~ predict(noSNMS3way), col = variety,
     data = fert2, subset = variety!='SNMS')
plot(resid(noSNMS3way) ~ fitted(noSNMS3way), col = variety,
     data = fert2, subset = variety!='SNMS')
legend('topright', pch = 1, col = 1:5,
       legend = levels(fert2$variety))
@
